15yrsChange

### 2024-04-05 doc/15yrsChange.Rmd
update methods pt. 2

In addition to our consideration of the entire forest plot, we focus further on three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams), with varying levels of exposure to non-indigenous insects and pathogens and white-tailed deer (Fig. 1).  The first focal area we define as“Low deer, low canopy vulnerability”: A 4-ha portion of upland forest in the SE quarter of the plot (Fig. 1a), fenced in 1990 and maintained with only occasional deer presence over the past 24 years. This area has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens [@holm_interactive_2013]. The second portion is a “high deer, low canopy vulnerability” 16.3 ha upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig. 1b). The third section of the plot has high deer, high canopy vulnerability: corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens. We defined upland forest according to topographic wetness index, which was calculated for the forest plot by @mcgregor_tree_2021a. We defined vulnerability to non-indigenous insects and pathogens based on the abundance of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment (2008). The following canopy species were identified as vulnerable: ash (*Fraxinus* spp.), elm (*Ulmus* spp.), butternut (*Juglans cinerea*), and American chestnut (*Castanea dentata*) [@anderson-teixeira_longterm_2021].

>>>>

In addition to our consideration of the entire forest plot, we focus further on three upland forest portions of the plot (i.e., excluding low-lying areas around streams), with varying levels of exposure to non-indigenous insects and pathogens and white-tailed deer (Fig. 1).  The first portion we define as “Low deer, low canopy vulnerability”: a 4-ha portion of upland forest in the SE quarter of the plot (Fig. **1a**), fenced in 1990 and maintained with only occasional deer presence over the past 24 years. This portion has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens [@holm_interactive_2013]. The second focal area is “high deer, low canopy vulnerability”: 16.3 ha of upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig. **1b**). The third portion of the plot is "high deer, high canopy vulnerability": corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens. We defined upland forest according to topographic wetness index, which was calculated for the forest plot by [@mcgregor_tree_2021a]. We defined vulnerability to non-indigenous insects and pathogens based on the abundance (>20% of tree basal area) of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment in 2008. The following canopy species were identified as vulnerable: ash (*Fraxinus* spp.), elm (*Ulmus* spp.), butternut (*Juglans cinerea*), and American chestnut (*Castanea dentata*) [@anderson-teixeira_longterm_2021].
